---
output: github_document
editor_options: 
  markdown: 

    wrap: sentence
---

# Excersises

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  comment = "#>", 
  fig.path = "man/figures/README-", 
  out.width = "90%", 
  dpi = 300, 
  message = FALSE
)

```

Part of this workshop has been adapted from (Thomas Lin Pedersen) GitHub repository: [link](https://github.com/thomasp85/ggplot2_workshop)

Link to YouTube video: <https://www.youtube.com/watch?v=h29g21z0a68>

Plenty of examples for basic ggplot2 package here: [website](http://www.sthda.com/english/wiki/be-awesome-in-ggplot2-a-practical-guide-to-be-highly-effective-r-software-and-data-visualization)

145+ extensions to add to ggplot2: <https://exts.ggplot2.tidyverse.org/gallery/>

**It's dangerous to go alone! Take this [cheatsheet](https://rstudio.github.io/cheatsheets/data-visualization.pdf)**

### Installing the required packages:

Main packages used for exercises

```{r, message=FALSE}
packages <- c('ggplot2', 'dplyr', 'grid', 'gridExtra', 'tidyr', 'reshape2')

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Packages loading
invisible(lapply(packages, library, character.only = TRUE))
```

Optional packages used to show some plot examples (Only needed for plot examples in "GGplot_workshop.Rmd")

```{r, message=FALSE}
# Packages loading
invisible(lapply(packages, library, character.only = TRUE))

# Optional (for some examples):
optional_packages <- c(
  'maps',
  'ggrepel',
  'ggraph',
  'igraph',
  'tidygraph',
  'GGally',
  'ggtree',
  'wordcloud2',
  'gt',
  'RColorBrewer'
)

# This package is installed from github
if (! "chorddiag" %in% installed.packages()) {
  devtools::install_github("mattflor/chorddiag")
}

## Optional package: treeio is available in Bioconductor
if (! "treeio" %in% installed.packages()) {
  install.packages("BiocManager")
  BiocManager::install("treeio")
}

# Install packages not yet installed
installed_packages <- optional_packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(optional_packages[!installed_packages])
}

# Packages loading
invisible(lapply(optional_packages, library, character.only = TRUE))
```

### Reshaping data

A dataset can be written in two different formats: wide and long.
A wide format has values that do not repeat in the first column.
A long format has values that do repeat in the first column.
Iris data is presented in wide format.
In some cases ggplot will require long format in order to plot the data correctly.
To transform from wide to long we have several options:

```{r}
data(iris)
head(iris)
```

#### gather() from tidyr

```{r}
# gather() from tidyr
head(gather(iris,                      # Data object
            feature,                   # Name of new key column (made from names of data columns)
            value,                     # Name of new value column
            Sepal.Length:Petal.Width,  # Names of source columns that contain values
            factor_key=TRUE))          # Treat the new key column as a factor (instead of character vector)


```

#### melt() from reshape2

```{r}
# melt() from reshape2
head(
  melt(iris, id.vars=c("Species"))  # Specify id.vars: the variables to keep but not split apart on
)

# There are options for melt that can make the output a little easier to work with:
head(
  melt(iris,
       id.vars=c("Species"),                            # ID variables - all the variables to keep but not split apart on
       measure.vars=c("Sepal.Length", "Sepal.Width",    # The source columns
                      "Petal.Length", "Petal.Width"),
       variable.name="variable",                        # Name of the destination columns that will identify the original columns that the measurement came from
       value.name="value")
)

```

To transform from long to wide you can use spread()

# Basic plot with ggplot and mapping

The main function of ggplot is the `ggplot()` function.
It has two main arguments that you need to be aware of:

-   **Data**: the (default) dataset to use for the plot

-   **Mapping**: the aesthetic mappings

When calling `ggplot()` we typically don’t call these arguments by their name and instead simply supply them in order.

### Data

The data argument obviously needs a dataset and this should usually come in the form of a data.frame or a tibble.
Let’s starts with a very simple data set, iris, that contains 4 measures + species name for each plant.

### Mapping

The mapping argument should be supplied with a call to the `aes()` function that lists the mappings between variables in your ta and aesthetics of the plot.
Which aesthetics you choose to map to depends on what geoms you intend to use.
But for now, let’s assume that we want to map "Petal.Length" to the x axis, and "Petal.Width" to the y axis.
To do so, we simply use `aes(x = Petal.Length, y = Petal.Width)` as the input to the mapping argument.
Because the first two arguments of aes() are x and y, we can do without calling them explicitly.

```{r}
ggplot(data = iris,
       mapping = aes(x = Petal.Length,
                     y = Petal.Width))
```

Why does this result in a blank canvas?
After all, we did tell ggplot to map Petal.Length to the x axis and Petal.Width to he y axis?
The answer is that we still haven’t told ggplot2 which layers we want to use.

# Layers

To add new layers to the main function we only need to 'add' (`+`) the code for the new layer.
The aesthetic mapping is not enough by itself—we also need a layer to display our data.
In ggplot2, we accomplish this by adding (stacking) layers to the output of ggplot().
There are several ways to construct layers in ggplot, but the simplest and most common approach is to use the functions named `geom_*()` (such as `geom_point()`).

### Aesthetics

Aesthetics mapped in ggplot() will be shared between all layers.
Each aesthetic can have its own mapping.
You can look at the documentation of the `geom_*()` function using the help function in R

`?geom_point()`

Under Aesthetics we can see that this layer understands the following aesthetics:

-   x
-   y
-   alpha
-   colour
-   fill
-   group
-   shape
-   size
-   stroke

#### Exercises:

1)  Generate a dotplot for Petal.Length \~ Petal.Width

```{r}
```

2)  Add a regression line (geom_smooth() with `formula = y ~ x` and `method='lm'`

```{r}
```

3)  Using the previous plot, map the speceies data to the color of the dots. Try first to do the mapping on the `geom_point` function and then mapping it on the main function `ggplot()`

```{r}
```

4)  Change size and shape of dots

```{r}
```

5)  Add a rectangle (`geom_rect()`) and fill it in green (`?geom_rect`)

```{r}
```

# Geoms

#### Exercises:

1)  Plot Species names instead of dots using geom_text. You will need to map Species names to `label`

```{r}
```

2)  Generate a boxplot or violinplot for Petal.Length data with one box for each species

```{r}
```

3)  Repeat the boxplot but now try to have all variables (Petal.Length, Petal.Width, Sepal.Length, Sepal.Width) in the same lot (\*You need the long format to do this plot)

```{r}
```

4)  Plot the distribution of Petal.Length by species (`geom_density()`)

```{r}
```

5)  Add the bar plot on top of the previous density plot. Use `position="dodge"` to plot the bars side by side

```{r}
```

# Stats

To know which statistics are calculated on each stat\_\* function check the section 'Computed variables' on its help page (`?stat_count`).

We can access this calculated values using after_stat()

#### Exercises:

1)  Generate a barplot showing car counts per classes using mpg dataset. Show counts in percentage of the total. This will require a new mapping in the geom

```{r}
ggplot(mpg, aes(x=class)) +
  stat_count()
# is the same than
ggplot(mpg, aes(x=class)) +
  geom_bar()
```

`geom_bar()` uses the default values from stat_count to generate the plot.
We can access the statistics calculated by stat_count using after_stat()

```{r}
ggplot(mpg, aes(x=class)) +
  geom_bar( aes( y = after_stat(100*count/sum(count)) ))
```

#### More examples

```{r}
# Add significance level
library("ggpubr")
ggplot(iris, aes(x=Species, y=Petal.Length, color=Species, fill=Species)) +
  geom_boxplot(alpha=.3) +
  stat_compare_means(label.y = 10) +
  stat_compare_means(comparisons = list(c(1,2), c(1,3), c(2,3)),
                     label = "p.signif") + theme_bw()

```

More examples with stat\_'geom' and stat_summary here: <https://ggplot2tutor.com/tutorials/summary_statistics>

# Multiplots

`facet_grid()` allows us to split data in different panels according to specific variables.
To do so it is recommended to use the long format.
This function will split (map) the data into `rows ~ columns`.
If we only want to split data into rows then we add a "." for columns, like this: `rows ~ .`

#### Exercises:

1)  Plot the distribution of each measure (`Sepal.Length`, `Sepal.Width`, ...) in different rows, map Species to colors

```{r}
```

2)  Now split the previous plot by placing each Species in different columns (keep variables in different row)

```{r}
```

3)  Make the plots easier to read by adjusting the Y scale in each row (check `?facet_grid` to find the correct option) facet_grid draws 1 plot on a canvas.

```{r}
```

We can draw different plots on 1 canvas using other libraries such as grid.arrange (libraries grid & gridExtra) or patchwork.
Shape of grid can be specified by a `layout_matrix`.
We need to pass a list of ggplot objects (or 'grobs') as input.

A graphical object (“grob”) is a description of a graphical item.
These basic classes provide default behavior for validating, drawing, and modifying graphical objects.

#### Exercises 2:

```{r}
# function to extract legend from plot
get_only_legend <- function(plot) {
  plot_table <- ggplot_gtable(ggplot_build(plot))
  legend_plot <- which(sapply(plot_table$grobs, function(x) x$name) == "guide-box")
  legend <- plot_table$grobs[[legend_plot]]
  return(legend)
}
```

1)  Save 2 ggplots into 2 different grob objects

```{r}
p1 <- ggplot(...)
p2 <- ggplot(...)
```

2)  Use grid.arrange() to plot the previous 2 objects together, either in 2 columns or in two rows

```{r}
```

3)  Draw the 2 plots with the same size and without legend ( `... + theme(legend.position = "none")` ) and make them share the same legend (Hint: use the function get_only_legend() described above to extract the legend as a grob)

```{r}
```

# Scales & Coordinates

Coordinates allows us to transform mappings.
For example:

-   Cartesian coordinates system allows us to zoom in and out of a plot

-   Polar coordinate system interprets x and y as angles and radius

    example: `p3 + scale_y_reverse()`

The commands for scales follow the pattern `scale_<aesthetic>_<mehtod>()`

#### Exercises:

1.1) Using `Sepal.Length` draw a bar plot and fill the bars by species

```{r}
```

1.2) Scale x axis into bins

```{r}
```

1.3) zoom in by scaling the y axis (`scale_y_continuous`) and set the limits to show only data below 10 counts

```{r}
```

1.4) repeat the zoom in but this time using `coord_cartesian()`.
Compare results with previous step

```{r}
```

2.1) Using the following exponential data transform the axis so dots follow a straight line (scale Y axis to log10)

```{r}
data <- data.frame(x=1:100, y=10**c(1:100))
```

3.1) Use the following data to create a bar chart with only one stacked bar (`stat="identity"`, or use `geom_col()`), and fill by group (hint: You can always map "" to a aesthetic)

```{r}
data <- data.frame(
  group=LETTERS[1:5],
  value=c(13,7,9,21,2)
)
```

3.2) Transform the y axis of previous plot into angles using `coord_polar(...)`

```{r}
```

```{r}
# BE CAREFUL WITH TRANSFORMATIONS!!
movies <- ggplot2movies::movies
mov <- movies[sample(nrow(movies), 1000), ]
mov
m1 <- ggplot(mov, aes(x = factor(round(rating)), y = votes)) +
  geom_point() +
  stat_summary(fun.data = "mean_cl_boot",
               geom = "crossbar",
               colour = "red", width = 0.3) +
  xlab("rating")
# Transforming the scale means the data are transformed first,
# after which statistics are computed:
p1 <- m1 + scale_y_log10(breaks=c(10000, 20000, 30000, 40000, 50000, 100000, 150000))
p1
# Transforming the coordinate system occurs after the
# statistic has been computed. This means we're calculating the summary on the raw data
# and stretching the geoms onto the log scale.
# Compare the widths of the standard errors.
p2 <- m1 + coord_trans(y="log10") + scale_y_continuous(breaks=c(10000, 20000, 30000, 40000, 50000, 100000, 150000))
p2

grid.arrange(m1, p1, p2, ncol=3)

```

# Colors & themes

Manual control of color/fill scales The comands follow the pattern `scale_<aesthetic>_<mehtod>()` (`scale_fill_distiller()` - to provide a ColorBrewer palette `scale_color_viridis()` - to use Viridis).
Or you can use `scale_fill_manual()` to provide manually a list of colors to fill.
Do not forget `discrete=FALSE` for a continuous variable.

Color names for ggplot2: <http://sape.inf.usi.ch/quick-reference/ggplot2/colour>

```{r}

# Dummy data
x <- LETTERS[1:20]
y <- paste0("var", seq(1,20))
data <- expand.grid(X=x, Y=y)
data$Z <- runif(400, 0, 5)
```

#### Exercises:

1)  Using the data above draw a heatmap (`geom_tile()`)

```{r}
```

2)  Change the filling using different scale\_\*

```{r}
```

3)  Use iris data to draw again a box plot and change color and fill manually

```{r}
```

Any other aesthetic in the plot can be modify using theme.
Inside theme(...) you can control aesthetics for panel, plot, axis, legend, etc.
Each aesthetic is defined by an `element_'shape'` constructor (text, line, blank, etc; e.g. `ggplot() + theme(axis.text = element_text(size=12))` to control the fontsize of (both) axis texts).
There are some predefined themes in ggplot such as `theme_bw()`.
`theme_minimal()`, `theme_classic()`, etc.

#### Exercises 2:

1)  Modify the aesthetics of your last plot to make it more appealing

```{r}
```

# Annotation

This function adds geoms to a plot, but unlike a typical geom function, the properties of annotations are not mapped from variables of a data frame, but are instead passed in as vectors.
This is useful for adding small annotations (such as text labels)

#### Exercises:

Using the dot plot from the first examples...

```{r}
p <- ggplot(iris, aes(Petal.Length, Petal.Width, color=Species)) +
  geom_point(size=3) +
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth",
              alpha=.2)
p
```

1)  Annotate the names of the species inside the plot using annotate()

```{r}
p + annotate(geom = "text", x=1.5, y=0.75, label = "Setosa", col="red") +
  annotate(geom = "text", x=4,   y=0.75, label = "Versicolor", col="darkgreen") +
  annotate(geom = "text", x=6,   y=1.25, label = "Virginica", col="blue")
```

2)  Highlight dots of setosa by annotating a rectangle that cover all dots

```{r}
p + annotate("rect", xmin = 0.75, xmax = 2, ymin = 0, ymax = 0.75, alpha = .3, fill="yellow")
```
